Racial and Ethnic Differences in ETI: Outcomes

### 2024-05-12 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti.qmd
reorganized race and ETI files

```{r data cleaning}
#| include: false
# source("/Users/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_race_eti_analysis_dataset.R")
load("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/analysis_dataset.RData")
```

>>>>

```{r data cleaning}
#| include: false
# source("/Users/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_race_eti_prescription_dataset.R")
load("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/prescription_dataset.RData")
```

### 2024-05-12 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti_outcomes.qmd
reorganized race and ETI files

```{r data setup}
#| include: false
# Import
encounter <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_encountersMerged_Del1.csv", na.strings = "")
annual <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_Annualized_Del1.csv", na.strings = "")
demo <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_DemogCFDiag_Del1.csv", na.strings = "")
# Race variable
races <- list(
  "Race1" = "White",
  "Race2" = "Black or African American",
  "Race3" = "American Indian or Alaska Native",
  "Race4" = "Asian",
  "Race5" = "Native Hawaiian or Other Pacific Islander",
  "Race6" = "Some other race"
)
demo$Race <- apply(demo[, paste0("Race", 1:6)], 1, function(r) {
  w <- which(r == 1)
  if (length(w) > 1 | 6 %in% w) {
    return("Mixed/Other/Unknown Race")
  } else {
    return(as.character(races[w]))
  }
})
demo$Race[demo$Hispanicrace == 1] <- "Hispanic or Latino"
demo$Race[is.na(demo$Hispanicrace)] <- "Mixed/Other/Unknown Race"
demo$Race <- factor(demo$Race,
  levels = c(
    "White", "Black or African American", "Hispanic or Latino", "Asian",
    "American Indian or Alaska Native",
    "Native Hawaiian or Other Pacific Islander", "Mixed/Other/Unknown Race"
  ),
  labels = c(
    "Non-Hispanic White", "Black or African American", "Hispanic or Latino", "Asian",
    "Other", "Other", "Other"
  )
)
# Insurance
ins_vars <- c(paste0("insurance", 1:7), "noinsurance")
ins <- c(
  "Health Insurance Policy (e.g. Private Insurance)", "Medicare", "Medicaid",
  "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
  "TriCare or other military health plan",
  "Indian Health Service", "Other", "No Insurance"
)
annual$Insurance <- apply(annual[, ins_vars], 1, function(r) {
  w <- which(r == 1)
  if (length(w) == 0) {
    return(NA)
  } else if (length(w) > 1) {
    return("More than one")
  } else {
    return(ins[w])
  }
})
annual$Insurance <- factor(annual$Insurance,
  levels = c(
    "Health Insurance Policy (e.g. Private Insurance)",
    "Indian Health Service",
    "Medicaid",
    "Medicare",
    "More than one",
    "No Insurance",
    "Other",
    "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
    "TriCare or other military health plan"
  ),
  labels = c(
    "Private", "Public", "Public", "Public", "More Than One", "No Insurance",
    "No Insurance", "Public", "Tricare"
  )
)
# Merge
encounter <- left_join(
  encounter,
  demo %>% select(eDWID, Race, Modulator_trikafta_first_date)
)
encounter <- left_join(encounter,
  annual %>% select(eDWID, ReviewYear, Insurance),
  by = join_by(eDWID, reviewyear == ReviewYear)
)
# Age group
encounter$age_group <- cut(encounter$encounterage, breaks = c(6, 12, 18, 30, Inf), right = F)
# Days between encounter and first ETI
encounter$Modulator_trikafta_first_date <-
  mdy(encounter$Modulator_trikafta_first_date)
encounter$encounterdate <- mdy(encounter$encounterdate)
encounter$Days <- as.numeric(encounter$encounterdate - encounter$Modulator_trikafta_first_date)
# Separate cleaned set
cleaned <- encounter %>% filter(gli_fev1_ppred_rn <= 200, !is.na(Days))
# List of outcomes
continuous_outcomes <- c(
  "gli_fev1_ppred_rn", "gli_fvc_ppred_rn", "gli_fev1fvc_ppred_rn",
  "bmipercentile", "heightpercentile", "weightpercentile"
)
binary_outcomes <- c(
  "staphylococcus_aureus", "haemophilus_influenzae", "pseudomonasaeruginosa",
  "burkho_complex"
)
```

@@

```{r}
#| cache: true
set.seed(1017)
model <- list(
  gli_fev1_ppred_rn ~ 1, # intercept
  gli_fev1_ppred_rn ~ 1 ~ 0 + Days # joined slope
)
fit <- mcp(model, data = cleaned, chains = 4, cores = 4)
```

```{r}
summary(fit)
plot(fit)
```

<!-- ## Frequentist models -->

<!-- First, a Gaussian mixed model is fit with random intercept for patient ID and random slope for each patient. Then, the `segmented.lme` function is applied to this mixed model, with 0 as the starting value for finding the optimal breakpoint. The changepoint is also treated as a random effect and allowed to vary within a person. -->

<!-- ### FEV1 -->

<!-- ```{r} -->
<!-- #| cache: true -->
<!-- # Fit initial model -->
<!-- ri_rs_mod_fev1 <- lme(gli_fev1_ppred_rn ~ Race * Days + age_group + Insurance, -->
<!--   random = ~ 1 + Days | eDWID, data = cleaned, na.action = na.omit -->
<!-- ) -->
<!-- # Find change point -->
<!-- ri_rs_mod_seg_fev <- segmented.lme(ri_rs_mod_fev1, ~Days, psi = 0, -->
<!--   random = list(eDWID = pdDiag(~ 1 + Days + U + G0)), # ind. random eff. (in the changepoint too) -->
<!--   control = seg.control(n.boot = 0, display = F) -->
<!-- ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- plot(ri_rs_mod_seg_fev1) -->
<!-- ``` -->

>>>>

```{r data setup}
#| include: false
# Import
encounter <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_encountersMerged_Del1.csv", na.strings = "")
annual <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_Annualized_Del1.csv", na.strings = "")
demo <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_DemogCFDiag_Del1.csv", na.strings = "")
# Race variable
races <- list(
  "Race1" = "White",
  "Race2" = "Black or African American",
  "Race3" = "American Indian or Alaska Native",
  "Race4" = "Asian",
  "Race5" = "Native Hawaiian or Other Pacific Islander",
  "Race6" = "Some other race"
)
demo$Race <- apply(demo[, paste0("Race", 1:6)], 1, function(r) {
  w <- which(r == 1)
  if (length(w) > 1 | 6 %in% w) {
    return("Mixed/Other/Unknown Race")
  } else {
    return(as.character(races[w]))
  }
})
demo$Race[demo$Hispanicrace == 1] <- "Hispanic or Latino"
demo$Race[is.na(demo$Hispanicrace)] <- "Mixed/Other/Unknown Race"
demo$Race <- factor(demo$Race,
  levels = c(
    "White", "Black or African American", "Hispanic or Latino", "Asian",
    "American Indian or Alaska Native",
    "Native Hawaiian or Other Pacific Islander", "Mixed/Other/Unknown Race"
  ),
  labels = c(
    "Non-Hispanic White", "Black or African American", "Hispanic or Latino", "Asian",
    "Other", "Other", "Other"
  )
)
# Insurance
ins_vars <- c(paste0("insurance", 1:7), "noinsurance")
ins <- c(
  "Health Insurance Policy (e.g. Private Insurance)", "Medicare", "Medicaid",
  "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
  "TriCare or other military health plan",
  "Indian Health Service", "Other", "No Insurance"
)
annual$Insurance <- apply(annual[, ins_vars], 1, function(r) {
  w <- which(r == 1)
  if (length(w) == 0) {
    return(NA)
  } else if (length(w) > 1) {
    return("More than one")
  } else {
    return(ins[w])
  }
})
annual$Insurance <- factor(annual$Insurance,
  levels = c(
    "Health Insurance Policy (e.g. Private Insurance)",
    "Indian Health Service",
    "Medicaid",
    "Medicare",
    "More than one",
    "No Insurance",
    "Other",
    "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
    "TriCare or other military health plan"
  ),
  labels = c(
    "Private", "Public", "Public", "Public", "More Than One", "No Insurance",
    "No Insurance", "Public", "Tricare"
  )
)
# Merge
encounter <- left_join(
  encounter,
  demo %>% select(eDWID, Race, Modulator_trikafta_first_date)
)
encounter <- left_join(encounter,
  annual %>% select(eDWID, ReviewYear, Insurance),
  by = join_by(eDWID, reviewyear == ReviewYear)
)
# Age group
encounter$age_group <- cut(encounter$encounterage, breaks = c(6, 12, 18, 30, Inf), right = F)
# Days between encounter and first ETI
encounter$Modulator_trikafta_first_date <-
  mdy(encounter$Modulator_trikafta_first_date)
encounter$encounterdate <- mdy(encounter$encounterdate)
encounter$Days <- as.numeric(encounter$encounterdate - encounter$Modulator_trikafta_first_date)
# Separate cleaned set
cleaned <- encounter %>% filter(gli_fev1_ppred_rn <= 200, !is.na(Days))
# List of outcomes
continuous_outcomes <- c(
  "gli_fev1_ppred_rn", "gli_fvc_ppred_rn", "gli_fev1fvc_ppred_rn",
  "bmipercentile", "heightpercentile", "weightpercentile"
)
binary_outcomes <- c(
  "staphylococcus_aureus", "haemophilus_influenzae", "pseudomonasaeruginosa",
  "burkho_complex"
)
# Save dataset for mcp model script
save(encounter, file = "./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/mcp_dataset.RData")
```

@@

## Frequentist models

First, a Gaussian mixed model is fit with random intercept for patient ID and random slope for each patient. Then, the `segmented.lme` function is applied to this mixed model, with 0 as the starting value for finding the optimal breakpoint. The changepoint is also treated as a random effect and allowed to vary within a person.

```{r}
#| cache: true
#| eval: false
# Fit initial model
ri_rs_mod_fev1 <- lme(gli_fev1_ppred_rn ~ Race * Days + age_group + Insurance,
  random = ~ 1 + Days | eDWID, data = cleaned, na.action = na.omit
)
# Find change point
ri_rs_mod_seg_fev <- segmented.lme(ri_rs_mod_fev1, ~Days, psi = 0,
  random = list(eDWID = pdDiag(~ 1 + Days + U + G0)), # ind. random eff. (in the changepoint too)
  control = seg.control(n.boot = 0, display = F)
)
```

```{r}

```
